Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
bugfix: memory consumption of naïve layout
Reviewer A:
  p9, l372: “assuming that booleans are stored in the smallest
  addressable memory size, a byte” vous êtes surprenamment généreux,
  la réalité est généralement bien plus en faveur de votre argument,
  les valeurs sont alignés sur des mots, donc un booléen prend
  typiquement 4 à 8 octets.

En effet, quelle bévue !
About constant-time vs. abstractions
Reviewer A:
  Je m'attendais un aussi à voir une algébraisation, ou du moins une
  abstraction, qui garantie que l'implémentation n'utilise pas
  d'opérateur de contrôle, mais il ne me sembl

@@

If we were to adopt \coqe!state! as our actual representation, the
memory usage of the resulting program would be quite poor. It would take
at least 128 bytes to store that many bits (assuming that booleans are
stored in the smallest addressable memory size, a byte). This would be
a terrible waste of memory bandwidth and register usage.

@@

We must therefore look toward adopting a \emph{packed} representation,
grouping individual booleans into a single machine word. Doing so
would improve memory density. However, we have to be very careful that
we can still efficiently compute over this packed representation. As
it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
now: we will have to wait until Section~\ref{sec:fixslicing}, to have
our cake and eat it too (using the \emph{same} data representation).

>>>>

Remarkably, reading off the types of \coqe!subcells_!,
\coqe!add_round_key_!, \coqe!shiftrows_! and \coqe!mixcolumns_!, we
can argue that this specification supports an implementation immune to
time-based side-channel attacks. This boils to justifying of the
absence of control-flow operation depending on cryptographic secrets
(the plaintext \coqe!s! and, even more crucially, the round keys in
\coqe!constkey!). It is obvious for \coqe!shiftrows_!, which is
defined parametrically over any type \coqe!A!. If we can guarantee
that the Boolean algebra operations can be implemented in
constant-time (which is the objective of the next Section), then we
get that the other 3 operations and, therefore, their composition is
constant-time too.

@@

If we were to adopt \coqe!state! as our actual representation, the
memory usage of the resulting program would be quite poor. It would
take at least 512 bytes to store these 128 Booleans (assuming that each
boolean is stored in the smallest addressable memory size, a 4-bytes
word). This would be a terrible waste of memory bandwidth and register
usage.

@@

We must therefore look toward adopting a \emph{packed} representation,
grouping individual booleans into a single machine word. Doing so
would improve memory density. This would also lead to an
implementation of the Boolean algebra that operates uniformly over
whole machine words, using machine instructions to process an entire
slice in a set number of CPU cycles. This is the fundamental
assumption behind bitslicing~\citep{biham:DES}: applying, say, a
logical \coqe!AND! machine instruction over two registers (each
representing a slice of the cipher state) takes one CPU cycle and
therefore takes the same amount of time, independently of the values
stored in the registers\footnote{A careful reader will have noted the
non sequitur: taking one CPU cycle does not actually imply
taking identitical (physical) execution time~\citep{wang:hertzbleed}. In very specific circumstances where the cryptographic primitive acts as a signal amplifier, this can actually be exploited: we do not account for this sort of second-order effect in the present work.}.

%% However, we have to be very careful that
%% we can still efficiently compute over this packed representation. As
%% it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
%% now: we will have to wait until Section~\ref{sec:fixslicing}, to have
%% our cake and eat it too (using the \emph{same} data representation).
